Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-07 paper2.qmd
section:6; process:write

Among the many potential directions for future research--both current and new--three avenues deserve particular attention due to their direct impact on the reliability and validity of CJ trait estimates, as well as on the accuracy of statistical inferences. The following sections outline these avenues and explain how our approach facilitates their investigation.

@@

<!-- 1. The overlooked role of the sampling and comparison mechanisms in CJ -->
Although sampling and comparison mechanisms are central to modern CJ assessments, it is striking that most of the CJ literature has examined them only within a limited scope. Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022]. While these studies offer valuable insights, they often overlook the broader role these mechanisms play within the CJ system. As this oversight likely stems from a broader lack of conceptual clarity about how these mechanisms function within the system, this study decided to formally incorporate these mechanisms into CJ's conceptual model.

<!-- 2. The missing data perspective on sampling and comparison mechanisms -->
The explicit integration of the sampling and comparison mechanisms offers a new perspective on how these mechanisms shape the CJ process. In particular, it clarifies their role in the data-generating process by framing them as sources of missing data--that is, as mechanisms that determine which observations are missing from the final data sample. This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these missing data mechanisms, their influence on CJ outcomes, and their implications for designing and assessing more complex assessments setups.

@@

<!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
However, many modern CJ applications rely on more complex assessment designs, in which the sampling and comparison mechanisms introduce more intricate forms of missingness — such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020]. A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015]. This pair selection process suggests that ACJ's comparison mechanism is outcome-dependent, potentially classifying the method as a generator of MNAR data. If this classification holds, the mixed findings on ACJ's capabilities become more comprehensible: some studies find that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others contend that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].

@@

<!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
Despite the growing notion that various stimulus-related factors can influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a]. This gap likely persists not due to a lack of interest or research but because researchers often rely on ad-hoc detection methods--such as 'misfit' statistics--that are poorly suited for the task [@Kelly_et_al_2022]. To address this limitation, this study treats judges' biases as an integral component of the CJ system from the outset. This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons, one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.

@@

<!-- 2. Relevant research question paved by recognizing judges' bias (part I) -->
The explicit integration of judges' bias into CJ's conceptual model then paves the way for investigating several relevant research questions. One key question is whether researchers can validly analyze CJ data under the assumption of "sample-free" trait calibration, specifically under the hypothesis that judges exhibit no systematic bias. This question is particularly relevant because many researchers still regard "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases. Another critical question, discussed in @sec-theory-issue1b, is whether training or expertise can help judges avoid focusing on irrelevant stimulus features--such as handwriting--over more central criteria like argumentative quality in writing assessments [@Kelly_et_al_2022]. Notably, exploring these questions may also provide insights into what it truly means to be an "expert" within the CJ context [@Kelly_et_al_2022].

@@

<!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
Moreover, since judges draw on a variety of intricate stimulus characteristics when evaluating complex, multidimensional traits [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], and these characteristics account for variation in judgment accuracy [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020; @Gijsen_et_al_2021], it is reasonable to expect that judges' assessments also vary with attributes such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022]. Prior studies support this view [@Bartholomew_et_al_2020a; @McMahon_et_al_2015]. Thus, building on the discussion in @sec-discussion_RA1, researchers could further explore how judges' selection influences the formation of a "shared consensus" and whether these attributes introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022]. Furthermore, if such attributes indeed undermine the assumption of "sample-freeness," researchers should explore ways to control their influence and determine how many judges--and how many judgments per judge--are needed to produce reliable trait estimates in these contexts. Additionally, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy. Our approach provides the necessary structure to rigorously investigate these questions.

@@

### The identification of 'misfitting' stimuli and judges {#sec-discussion_RA3}
<!-- recording: Sven and Tine 25.05.21; time: 00:31:45 - 00:32:20 -->
<!-- recording: Sven and Tine 25.06.04; time: 00:33:55 - 00:53:15 -->

<!-- without disentangling all the potential processes the concept of misfits is strange, because it is not defined from Thustonian theory. As it is how is treated in CJ literature, misfit are outliers. People have theories on what makes misfits, but the approach now allows to test some hypothesis that might generate outliers under the BTL model, but no longer considered outliers in our approach  -->

<!-- the approach makes it possible to have hypothesis about why judges are labeled as misfits.  -->

Although the CJ literature clearly defines *misfit* judges and stimuli, researchers have rarely examined how these observations relate to Thurstonian theory. In particular, they have not identified which elements of Thurstone's theory might explain the occurrence of misfits. This disconnect likely stems from the fact that CJ researchers derive misfit statistics from methods such as residual analysis and outlier detection rather than from Thurstonian principles [@Pollitt_2012a; @Pollitt_2012b; @Wu_et_al_2022]. They typically define *misfit judges* as those whose assessments diverge significantly from the "shared consensus" [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and *misfit stimuli* as those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012a; @Pollitt_2012b; @Goossens_et_al_2018]. Both definitions closely mirror the statistical concept of outliers--that is, observations that deviate markedly from the rest of the sample in which they occur [@Grubbs_1969]. However, this resemblance extends beyond these definitions. CJ researchers also identify misfits using standard outlier detection procedures, such as transforming BTL model residuals into statistics of interest and comparing them against a predefined threshold.

@@

[After identifying misfit observations, researchers often propose various hypotheses to explain their presence. Yet, without providing evidence for any of these explanations, they typically exclude these observations from analysis. This exclusion process raises three major concerns: one related to the reliability of the methods used to identify misfits, a second concerning the potential impact of their removal on trait estimates and inferences, and a third regarding the hypotheses used to explain the presence and exclusion of misfits.]{style="color:blue;"}

[First, regarding the reliability of identification methods, it is relevant to note that outlying observations can only be defined relative to a specific model [@McElreath_2020]. Consequently, methods based on models such as the BTL--which rest on strong assumptions--should be treated with caution, as these assumptions may prevent the model from accurately capturing the underlying data-generating process of the CJ system, thereby potentially compromising the identification of misfits.]{style="color:blue;"}

@@

[Second, concerning the impact of removing misfits, the statistical literature warns that ad-hoc outlier exclusion procedures risk discarding valuable information [@Miller_2023] and introducing bias into trait estimates. The direction and magnitude of this bias are often unpredictable, as they depend on which observations researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].]{style="color:blue;"}

@@

[Lastly, regarding the hypotheses explaining the presence and exclusion of misfits, it is important to note that a significant number of these hypotheses could not be tested within the BTL model due to its strong assumptions and limited flexibility. This is the reason why there is a lack of supporting evidence for misfit observations. In contrast, our approach enables researchers to test these hypotheses. For instance, researchers could examine whether misfit judges are simply those who exhibit (an outlying degree of) systematic bias or whether they display greater variability in their judgments compared to their peers. Similarly, in the case of misfit stimuli, researchers could investigate whether their status reflects an increased variability in their discriminal processes or stems from their classification as outlying stimuli.]{style="color:blue;"}

[Moreover, since outliers are not necessarily "bad data" [@McElreath_2020], our approach offers a principled alternative that retains misfits in the analysis without compromising trait estimation. This alternative involves adapting our proposed models into more sophisticated, robust measurement models [@Wainer_et_al_1978]--a broad class of procedures designed to reduce the sensitivity of parameter estimates to mild or moderate departures from model assumptions [@Everitt_et_al_2010].]{style="color:blue;"}

@@

Even after acquiring the necessary foundational knowledge, CJ researchers face two additional challenges when applying our approach to their specific research context. First, they must verify whether the conceptual model provides a *faithful* [^12] representation of the CJ system under study. Second, they need to assess whether the statistical translation of the model can accurately estimate the intended estimands (i.e., parameters) from empirical data. To ensure a faithful representation of the CJ system, we encourage researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets. While these models may prove adequate in some contexts, researchers cannot assume this apriori. Instead, they need to adapt these models to their specific CJ contexts, paying close attention to assessment design features and causal assumptions. As discussed in @sec-discussion_challenges, our approach facilitates this process by offering a transparent framework for articulating new assumptions and guiding the design of CJ assessments.

@@

Conversely, to evaluate the estimation capabilities of the statistical model, researchers must tackle the challenge of performing identification analysis. As outlined in @sec-theoretical, *identification analysis* determines whether a statistical model can accurately compute a given estimand (e.g., a parameter) based solely on its (causal) assumptions, independent of random variability [@Schuessler_et_al_2023]. Identification is crucial because it is a necessary condition for consistency. *Consistency* is the property of an estimator (e.g., a statistical model) whose estimates converge to the "true" value of an estimand as data size approaches infinity [@Everitt_et_al_2010]. Without identification, consistency is impossible--even with infinite, error-free data--and thus, meaningful inference from finite samples cannot be achieved [@Schuessler_et_al_2023].

Although performing identification analysis through formal derivations may seem like a natural next step, we argue that the complexity of the CJ system renders such an approach impractical at the outset. Instead, simulation-based methods, such as power analysis, provide a more practical and flexible alternative, enabling researchers to examine the consistency of estimates without relying on cumbersome mathematical proofs. Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research. Importantly, our approach supports both strategies by providing the probabilistic foundation for formal derivations and the necessary statistical structure for simulation-based methods.

>>>>

Among the many potential directions for future research, three avenues deserve particular attention due to their direct impact on the reliability and validity of CJ trait estimates, as well as on the accuracy of statistical inferences. The following sections outline these avenues and explain how our approach facilitates their investigation.

@@

<!-- 1. The overlooked role of the sampling and comparison mechanisms in CJ -->
Although sampling and comparison mechanisms are central to modern CJ assessments, it is striking that most CJ literature has examined them within a limited scope. Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022]. While these studies offer valuable insights, they also overlook the broader role that these mechanisms play within the CJ system. As this oversight likely stems from a more fundamental lack of conceptual clarity about how these mechanisms function within the system, this study decided to integrate these mechanisms into the conceptual model of CJ.

<!-- 2. The missing data perspective on sampling and comparison mechanisms -->
The explicit integration of the sampling and comparison mechanisms offers a new perspective on how these mechanisms shape the CJ process. Specifically, it clarifies their role as sources of missing data in CJ's data-generating process--that is, as mechanisms that determine which observations are missing from the final data sample. This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], allowing a more rigorous evaluation of existing claims about these missing data mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex assessments setups.

@@

<!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
However, many modern CJ applications rely on more complex assessment designs, in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020]. A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015]. This pair selection process suggests that ACJ's comparison mechanism is outcome-dependent, potentially classifying the method as a generator of MNAR data. If this classification holds, the mixed findings on ACJ's capabilities become more comprehensible: some studies find that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others contend that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].

@@

<!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
Despite the growing notion that various stimulus-related factors influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a]. This gap likely persists not due to a lack of interest or research but because researchers often rely on ad-hoc detection methods, such as 'misfit' statistics, that may not be well-suited for the task [@Kelly_et_al_2022]. To address this limitation, this study treats judges' biases as an integral component of the CJ system from the outset. This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons, one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.

@@

<!-- 2. Relevant research question paved by recognizing judges' bias (part I) -->
The explicit integration of judges' bias into CJ's conceptual model then paves the way for investigating several relevant research questions. One key question is whether researchers can validly analyze CJ data under the assumption of "sample-free" trait calibration, specifically under the hypothesis that judges exhibit no systematic bias. This question is particularly relevant because many researchers still regard "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases. Another critical question, discussed in @sec-theory-issue1b, is whether training or expertise can help judges avoid focusing on irrelevant stimulus features, such as handwriting, over more central criteria like argumentative quality in writing assessments [@Kelly_et_al_2022]. Notably, exploring these questions may also provide insights into what it truly means to be an "expert" within the CJ context [@Kelly_et_al_2022].

@@

<!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
Moreover, since judges rely on these stimulus-related factors when evaluating complex, multidimensional traits [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], and these factors account for variation in judgment accuracy [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020; @Gijsen_et_al_2021], it is reasonable to expect that assessments also vary according to judge-specific attributes such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022]. Prior studies support this view [@Bartholomew_et_al_2020a; @McMahon_et_al_2015]. Thus, building on the discussion in @sec-discussion_RA1, researchers could further explore how judges' selection influences the formation of a "shared consensus" and whether these attributes introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022]. Furthermore, if such attributes indeed undermine the assumption of "sample-freeness," it becomes essential to explore strategies for mitigating their influence and to determine how many judges (and how many judgments per judge) are needed to produce reliable trait estimates under these conditions. In addition, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy. As anticipated, our approach provides the necessary structure to investigate rigorously these questions.

@@

### The identification of 'misfitting' judges and stimuli {#sec-discussion_RA3}
<!-- recording: Sven and Tine 25.05.21; time: 00:31:45 - 00:32:20 -->
<!-- recording: Sven and Tine 25.06.04; time: 00:33:55 - 00:53:15 -->

Although the CJ literature clearly defines *misfit* judges and stimuli, CJ researchers have rarely examined how these observations relate to Thurstonian theory. In particular, they have not identified which elements of Thurstone's theory might account for the occurrence of misfits. This disconnect likely stems from the fact that CJ researchers derive misfit statistics from residual analysis and outlier detection methods rather than from Thurstonian principles. Specifically, *misfit judges* are typically defined as those whose assessments diverge significantly from the "shared consensus" [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], while *misfit stimuli* are those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012a; @Pollitt_2012b; @Goossens_et_al_2018]. Both definitions closely mirror the statistical concept of outliers--that is, observations that deviate markedly from the rest of the sample in which they occur [@Grubbs_1969]. But this resemblance extends beyond the definitions themselves, as CJ researchers often identify misfits using conventional outlier detection procedures, such as transforming BTL model residuals into diagnostic statistics and comparing them against predefined thresholds [@Pollitt_2012a; @Pollitt_2012b; @Wu_et_al_2022].

@@

Nevertheless, it is not the classification of misfits as outliers that raises concerns for trait measurement and inference, but rather the prevailing CJ practice of identifying them through these ad-hoc detection procedures and then excluding them from analysis, often without providing empirical evidence for the various hypotheses proposed to explain their occurrence in the first place. In this regard, it is essential to recognize that outliers can only be defined relative to a specific model [@McElreath_2020]. Thus, identification procedures based on models like the BTL, which relies on strong assumptions, should be approached with caution, as they may not be well-suited for the task [@Kelly_et_al_2022]. Specifically, because the BTL model may not always accurately reflect the underlying data-generating process of the CJ system (see @sec-theory-issues), detection procedures grounded in this model risk compromising the accurate identification of misfits. Moreover, the statistical literature cautions that excluding outliers risks discarding valuable information [@Miller_2023] and introducing bias into trait estimates. The direction and magnitude of this bias are often unpredictable, as they depend on which observations researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020]. Lastly, regarding the hypotheses proposed to explain the presence of misfits, it is evident that the BTL model lacks the flexibility needed to adequately test many of them, largely due to its strong assumptions.

@@

@@

In contrast, our approach provides a rigorous framework that enables the examination of several relevant hypotheses. For instance, researchers could investigate whether misfit judges are those who exhibit (an outlying degree of) systematic bias or greater variability in their judgments compared to their peers. Similarly, researchers might explore whether misfit stimuli exhibit more variable discriminal processes relative to other stimuli or if they are genuinely outlying cases. Moreover, since outliers are not necessarily "bad data" [@McElreath_2020], our approach also offers a principled alternative that retains misfits in the analysis without compromising trait estimation or inference. This alternative involves adapting our proposed models into robust measurement models [@McElreath_2020]--a broad class of procedures designed to reduce the sensitivity of parameter estimates to mild or moderate departures from model assumptions [@Everitt_et_al_2010]. This alternative also relates to a broader discussion in social science, which is: that when researchers face low predictive capacity from their models, they should not only search for "new" variables or alternative procedures but also consider employing more sophisticated measurement models [@Wainer_et_al_1978].

@@

Even after acquiring the necessary foundational knowledge, CJ researchers still face two additional challenges when applying our approach to their specific research context. First, they must verify whether the conceptual model provides a *faithful* [^12] representation of the CJ system under study. Second, they need to assess whether the statistical translation of the model can accurately estimate the intended estimands (i.e., parameters) from empirical data. To ensure a faithful representation of the CJ system, we encourage researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets. While these models may prove adequate in some contexts, researchers cannot assume this apriori. Instead, they need to adapt these models to their specific contexts, paying close attention to assessment design features and causal assumptions. As discussed in @sec-discussion_challenges, our approach facilitates this process by offering a transparent framework for articulating new assumptions and guiding the design of CJ assessments.

@@

Conversely, to evaluate the estimation capabilities of the statistical model, researchers must tackle the challenge of performing identification analysis. As outlined in @sec-theoretical, *identification analysis* determines whether a statistical model can accurately compute a given estimand (e.g., a parameter) based solely on its (causal) assumptions, independent of random variability [@Schuessler_et_al_2023]. Identification is crucial because it is a necessary condition for consistency. *Consistency* is the property of an estimator (e.g., a statistical model) whose estimates converge to the "true" value of an estimand as data size approaches infinity [@Everitt_et_al_2010]. Without identification, consistency is impossible--even with infinite, error-free data--and thus, meaningful inference from finite samples cannot be achieved [@Schuessler_et_al_2023]. Although performing identification analysis through formal derivations may seem like a natural next step, the complexity of the CJ system renders such an approach impractical at the outset. Instead, simulation-based methods, such as power analysis, provide a more practical and flexible alternative, enabling researchers to examine the consistency of estimates without relying on cumbersome mathematical proofs. Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research. Importantly, our approach supports both strategies by providing the probabilistic foundation for formal derivations and the necessary statistical structure for simulation-based methods.
